---
<<<<<<< ours
title: 'Uncorking Quality: A Data-Driven Exploration of Wine Chemistry'
=======
title: 'Uncorking Quality: Unsupervised Exploration of Wine Chemistry'
>>>>>>> theirs
author: "Alexis Bryant"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(cluster)
library(factoextra)
library(GGally)
<<<<<<< ours
<<<<<<< ours
library(ggplot2)
=======
library(nnet)

set.seed(123)
theme_set(theme_minimal())
>>>>>>> theirs
=======
library(mclustcomp)

set.seed(123)
theme_set(theme_minimal())
>>>>>>> theirs
```

# Introduction

<<<<<<< ours
<<<<<<< ours
This project uses a Kaggle dataset containing the results of chemical analyses of wines grown in the same region of Italy. Each record represents one wine sample with 13 laboratory-measured chemical attributes.The goal of this project is to analyze the data using an unsupervised structure to uncover natural groupings and structure within the chemical features using PCA and k-means clustering.
=======
This project analyzes the classic Kaggle wine dataset containing laboratory measurements of wines grown in the same region of Italy.
Each of the **178 samples** has **13 chemical attributes** and a cultivar label (`Class`) indicating one of three grape varieties.
The analysis addresses three questions:

1. *Which chemical measurements drive the largest variability in the wines?*
2. *Do natural clusters in the chemistry align with the three cultivars?*
3. *How accurately can we classify cultivars from chemistry alone?*
>>>>>>> theirs

To answer them, the report combines exploratory data analysis (EDA), **principal component analysis (PCA)** for dimensionality reduction, **k-means clustering**, and two classification approaches (**multinomial logistic regression** and **LASSO-regularized multinomial regression**).

# Data Loading and Codebook

```{r}
wine <- read_csv("wine_dataset.csv", show_col_types = FALSE)
=======
This project analyzes the classic Kaggle wine dataset containing laboratory measurements of wines grown in the same region of Italy.
Each of the **178 samples** has **13 chemical attributes** but *no cultivar label*—only an unlabeled numeric `target` column (0, 1, 2) that may correspond to three grape groups.
Because cultivar identities are unknown, the analysis emphasizes unsupervised structure rather than supervised prediction.
The study addresses three questions:

1. *Which chemical measurements drive the largest variability in the wines?*
2. *Do natural clusters emerge from chemistry alone, and how stable are they?*
3. *How do unsupervised clusters compare to the anonymous `target` codes provided in the file?*

To answer them, the report combines exploratory data analysis (EDA), **principal component analysis (PCA)** for dimensionality reduction, **k-means clustering**, **hierarchical clustering**, and internal cluster validation (silhouette width and the gap statistic).

# Data Loading and Codebook

```{r}
data_path <- file.path("..", "Data", "wine_dataset.csv")
stopifnot(file.exists(data_path))

wine_raw <- read_csv(data_path, show_col_types = FALSE)

wine <- wine_raw %>%
  rename(
    Alcohol = alcohol,
    Malic_Acid = malic_acid,
    Ash = ash,
    Alkalinity_of_Ash = alcalinity_of_ash,
    Magnesium = magnesium,
    Total_Phenols = total_phenols,
    Flavanoids = flavanoids,
    Nonflavanoid_Phenols = nonflavanoid_phenols,
    Proanthocyanins = proanthocyanins,
    Color_Intensity = color_intensity,
    Hue = hue,
    OD280_OD315 = `od280/od315_of_diluted_wines`,
    Proline = proline,
    Target = target
  )
>>>>>>> theirs

glimpse(wine)
summary(wine)
```

<<<<<<< ours
<<<<<<< ours
This dataset includes variables such as:
-Alcohol
-Malic Acid
-Ash
-Alcalinity_of_Ash
-Magnesium
-Total_Phenols
-Flavanoids
-Nonflavanoid_Phenols
-Proanthocyanins
-Color_Intensity
-Hue
-0D280/0D315
-Proline

 All are numeric, making this dataset suitable for PCA and k-means clustering.
=======
**Codebook**
>>>>>>> theirs

```{r}
codebook <- tribble(
  ~Variable, ~Description,
  "Alcohol", "Alcohol content (% by volume)",
  "Malic_Acid", "Malic acid concentration (g/L)",
  "Ash", "Ash content (g/100 mL)",
  "Alcalinity_of_Ash", "Alkalinity of ash (mEq/L)",
  "Magnesium", "Magnesium content (mg/L)",
  "Total_Phenols", "Total phenolic compounds (g/L)",
  "Flavanoids", "Flavonoid content (g/L)",
  "Nonflavanoid_Phenols", "Non-flavonoid phenols (g/L)",
  "Proanthocyanins", "Proanthocyanins (g/L)",
  "Color_Intensity", "Color intensity (absorbance units)",
  "Hue", "Hue of the wine color (ratio)",
  "OD280_OD315", "Optical density ratio at 280/315 nm",
  "Proline", "Proline concentration (mg/L)",
  "Class", "Cultivar label (1, 2, or 3)"
)

knitr::kable(codebook, caption = "Wine attribute codebook")
```

# Data Processing
=======
**Codebook**
>>>>>>> theirs

<<<<<<< ours
Since the dataset consists of all numeric variables, we are going to standardize the data to ensure all variables contribute equally to the PCA and clustering steps.
```{r}
wine_scaled<- scale(wine)
```
We confirm that the scaling process centers the data around 0 and sets variance for each feature.
```{r}
<<<<<<< ours
apply(wine_scaled, 2, mean)
apply(wine_scaled, 2, sd)
```


# Principal Component Analysis (PCA)

PCA helps reduce dimensionality and visualize hidden structure. Each principal component represents a linear combination of features that capture the most variance.

```{r}

pca_res<-prcomp(wine_scaled, center = TRUE, scale. = TRUE)
summary(pca_res)
```


## Scree Plot

The scree plot shows the proportion of variance explained by each principal component. The first few components capture the majority of the datasets variability.

```{r}
factoextra::fviz_eig(pca_res, addlabels = TRUE, ylim = c(0, 50))
```
=======
```{r fig.width=10, fig.height=8}
wine_long <- wine %>%
  pivot_longer(cols = -Class, names_to = "Variable", values_to = "Value")

=======
codebook <- tribble(
  ~Variable, ~Description,
  "Alcohol", "Alcohol content (% by volume)",
  "Malic_Acid", "Malic acid concentration (g/L)",
  "Ash", "Ash content (g/100 mL)",
  "Alkalinity_of_Ash", "Alkalinity of ash (mEq/L)",
  "Magnesium", "Magnesium content (mg/L)",
  "Total_Phenols", "Total phenolic compounds (g/L)",
  "Flavanoids", "Flavonoid content (g/L)",
  "Nonflavanoid_Phenols", "Non-flavonoid phenols (g/L)",
  "Proanthocyanins", "Proanthocyanins (g/L)",
  "Color_Intensity", "Color intensity (absorbance units)",
  "Hue", "Hue of the wine color (ratio)",
  "OD280_OD315", "Optical density ratio at 280/315 nm",
  "Proline", "Proline concentration (mg/L)",
  "Target", "Anonymous numeric grouping code (0, 1, 2)"
)

knitr::kable(codebook, caption = "Wine attribute codebook")
```

# Exploratory Data Analysis

```{r fig.width=10, fig.height=8}
wine_long <- wine %>%
  pivot_longer(cols = -Target, names_to = "Variable", values_to = "Value")

>>>>>>> theirs
ggplot(wine_long, aes(x = Value)) +
  geom_histogram(bins = 25, fill = "#8C2D2D", color = "white") +
  facet_wrap(~Variable, scales = "free") +
  labs(title = "Distribution of Wine Chemistry Features", x = "Value", y = "Count")
```
<<<<<<< ours

Many features are right-skewed (e.g., **Proline**) while others are roughly symmetric.
This motivates standardizing variables before PCA, clustering, and regression.

```{r fig.width=7, fig.height=6}
GGally::ggcorr(wine %>% select(-Class), label = TRUE, label_round = 2, hjust = 0.75) +
  labs(title = "Correlation Matrix of Chemical Attributes")
```

Strong positive correlations among **Total_Phenols**, **Flavanoids**, and **OD280_OD315** suggest overlapping information that PCA can consolidate.

# Dimensionality Reduction with PCA

```{r fig.width=7, fig.height=5}
wine_scaled <- scale(select(wine, -Class))
pca <- prcomp(wine_scaled, center = TRUE, scale. = TRUE)

fviz_eig(pca) + labs(title = "Variance Explained by Principal Components")
```

```{r fig.width=7, fig.height=6}
fviz_pca_biplot(pca,
  habillage = wine$Class,
  addEllipses = TRUE,
  col.var = "gray30",
  repel = TRUE,
  labelsize = 3
) +
  labs(title = "PCA Biplot Colored by Cultivar")
```

The first two components capture most variability and visually separate the three cultivars.
**Flavanoids**, **Total_Phenols**, and **OD280_OD315** load heavily on PC1, highlighting their shared influence.
>>>>>>> theirs

## PCA Biplot

This plot shows how samples (points) and variables(arrows) relate in reduced dimensions. The direction and length of arrows indicate which features contribute most to each component.

<<<<<<< ours
```{r}
fviz_pca_biplot(pca_res, geom = "point") 
```
=======
```{r fig.width=7, fig.height=5}
set.seed(123)
km <- kmeans(wine_scaled, centers = 3, nstart = 25)
>>>>>>> theirs

fviz_cluster(km,
  data = wine_scaled,
  geom = "point",
  ellipse.type = "norm",
  ggtheme = theme_minimal(),
  main = "K-Means Clustering of Wines (k = 3)"
)
```

<<<<<<< ours
The first 2 PC's likely represent overall concentration and chemical richness of the wine samples.
Features such as Flavanoids, Color_Intensity, and Proline may have the strongest loadings, driving much of the variance.
=======
```{r}
sil_width <- silhouette(km$cluster, dist(wine_scaled))
mean_sil <- mean(sil_width[, 3])
mean_sil
```
>>>>>>> theirs

The average silhouette width (shown above) indicates well-separated clusters that align closely with the three cultivar labels.

<<<<<<< ours
# K-Means Clustering

## Determining Optimal Number of Clusters

```{r}
set.seed(123)
within_ss<-map_dbl(1:10, function(k){
  kmeans(wine_scaled, centers = k, nstart = 20)$tot.withinss
  })
tibble(k = 1:10, wss = within_ss) %>%
  ggplot(aes(k, wss)) + geom_line() + geom_point(size = 2, color = "blue") +
  labs(title = "Elbow Method for Optimal K", x = "Number of Clusters (K)", y = "Total Within Sum of Squares")
```

Based on the elbow point, we select K=3, consistent with prior studies on this dataset (which corresponds to 3 cultivars)


## Applying K-Means

```{r}
set.seed(123)
kmeans_res<- kmeans(wine_scaled, centers = 3, nstart = 25)
wine_clustered<- as_tibble(wine_scaled) %>% 
  mutate(Cluster = factor(kmeans_res$cluster))
```

## Visualizing Cluster Separation

```{r}
GGally::ggpairs(wine_clustered, columns = 1:3, aes(color = Cluster, alpha = 0.5))

factoextra::fviz_cluster(kmeans_res, data = wine_scaled, ellipse.type = "norm", geom = "point", main = "K-Means Clustering Results")
```
Interpretation:

The 3 clusters correspond to distinct groups of wines with differeing chemical balances. These may reflect natural subtypes based on cultivation or production style, even though the dataset does not explicitly include those labels
=======

Many features are right-skewed (e.g., **Proline**) while others are roughly symmetric.
This motivates standardizing variables before PCA and clustering.

```{r fig.width=7, fig.height=6}
GGally::ggcorr(wine %>% select(-Target), label = TRUE, label_round = 2, hjust = 0.75) +
  labs(title = "Correlation Matrix of Chemical Attributes")
```

Strong positive correlations among **Total_Phenols**, **Flavanoids**, and **OD280_OD315** suggest overlapping information that PCA can consolidate.

# Dimensionality Reduction with PCA

```{r fig.width=7, fig.height=5}
wine_scaled <- scale(select(wine, -Target))
pca <- prcomp(wine_scaled, center = TRUE, scale. = TRUE)

fviz_eig(pca) + labs(title = "Variance Explained by Principal Components")
```

```{r fig.width=7, fig.height=6}
fviz_pca_biplot(pca,
  col.ind = "gray40",
  col.var = "gray30",
  repel = TRUE,
  labelsize = 3
) +
  labs(title = "PCA Biplot (unsupervised)")
```

The first two components capture most variability and visually separate the wines along phenolic intensity.
**Flavanoids**, **Total_Phenols**, and **OD280_OD315** load heavily on PC1, highlighting their shared influence.

# Clustering

```{r fig.width=7, fig.height=5}
km <- kmeans(wine_scaled, centers = 3, nstart = 25)

fviz_cluster(km,
  data = wine_scaled,
  geom = "point",
  ellipse.type = "norm",
  ggtheme = theme_minimal(),
  main = "K-Means Clustering of Wines (k = 3)"
)
```

```{r}
sil_width <- silhouette(km$cluster, dist(wine_scaled))
mean_sil <- mean(sil_width[, 3])
mean_sil
```

The average silhouette width (shown above) indicates reasonably well-separated clusters even without cultivar labels.
>>>>>>> theirs

```{r fig.width=7, fig.height=5}
fviz_nbclust(wine_scaled, kmeans, method = "gap_stat", nstart = 25, nboot = 50) +
  labs(title = "Gap Statistic Suggests k = 3 Clusters")
```

<<<<<<< ours
# Correlation Analysis and Feature Relationships

Understanding how features correlate can provide insight into how certain chemicals vary together.

```{r}
library(corrplot)
cor_matrix<-cor(wine)

corrplot(cor_matrix, method = "color", tl.cex = 0.8, addCoef.col = "black", number.cex = 0.6, title = "Feature Correlation Heatmap: Wine Dataset")
```


Observations:

Flavanoids, Total_Phenols are strongly correlated, suggesting the co vary across wine samples.
Color_Intensity shows weaker correlations, indicating it captures unique variation.

=======
# Classification and Regression

## Multinomial Logistic Regression

```{r}
wine$Class <- as.factor(wine$Class)

split <- createDataPartition(wine$Class, p = 0.75, list = FALSE)
train <- wine[split, ]
test <- wine[-split, ]

mlogit_model <- multinom(Class ~ ., data = train)
mlogit_pred <- predict(mlogit_model, newdata = test)
mlogit_cm <- confusionMatrix(mlogit_pred, test$Class)
mlogit_cm
```

```{r}
mlogit_accuracy <- mlogit_cm$overall["Accuracy"]
mlogit_accuracy
```

The multinomial model achieves high accuracy, confirming that chemical measurements strongly predict cultivar.
>>>>>>> theirs

## LASSO-Regularized Multinomial Regression

```{r fig.width=7, fig.height=5}
x_train <- model.matrix(Class ~ . - 1, data = train)
y_train <- train$Class

cvfit <- cv.glmnet(x_train, y_train, family = "multinomial", alpha = 1)
plot(cvfit)
```

```{r}
x_test <- model.matrix(Class ~ . - 1, data = test)
lasso_pred <- predict(cvfit, newx = x_test, s = "lambda.min", type = "class") %>%
  factor(levels = levels(test$Class))
lasso_cm <- confusionMatrix(lasso_pred, test$Class)
lasso_cm
```

```{r}
lasso_accuracy <- lasso_cm$overall["Accuracy"]
lasso_accuracy
```

The LASSO model retains a subset of influential predictors while matching the performance of the unregularized multinomial model.

# Conclusions and Next Steps

- **Dimensionality reduction:** The first two PCs capture most variability and reveal clear cultivar separation driven by phenolic measurements.
- **Clustering:** K-means with three centers yields well-separated clusters (mean silhouette width above) that mirror the cultivar labels.
- **Classification:** Both multinomial and LASSO-regularized models achieve strong predictive accuracy, demonstrating that chemical profiles alone can identify cultivar.

**Future work** could explore random forests or gradient boosting for non-linear decision boundaries, incorporate external quality ratings, or test stability across vintages and regions.
=======
```{r fig.width=7, fig.height=5}
hc <- hclust(dist(wine_scaled), method = "ward.D2")
plot(hc, labels = FALSE, hang = -1, main = "Hierarchical Clustering Dendrogram")
rect.hclust(hc, k = 3, border = "#8C2D2D")
```

```{r}
hc_clusters <- cutree(hc, k = 3)

km_vs_hc <- table(kmeans = km$cluster, hclust = hc_clusters)
km_vs_hc
```

The gap statistic, dendrogram, and cross-tabulation show that both k-means and hierarchical clustering discover similar three-group structure, reinforcing the choice of **k = 3**.

```{r}
if ("Target" %in% names(wine)) {
  ari <- mclustcomp::ari(as.numeric(wine$Target) + 1, km$cluster)
  ari
}
```

Using the anonymous `Target` codes purely for post-hoc comparison, the adjusted Rand index (ARI) suggests strong agreement between the unsupervised clusters and the hidden grouping, despite not using the codes during clustering.

# Conclusions and Next Steps

- **Dimensionality reduction:** The first two PCs capture most variability and reveal separation driven by phenolic measurements.
- **Unsupervised clustering:** K-means and hierarchical clustering agree on a three-cluster solution with solid silhouette support.
- **Hidden-label check:** Post-hoc ARI comparison indicates that the anonymous `Target` codes align well with the discovered clusters, lending confidence that chemistry alone recovers meaningful groupings.

**Future work** could layer on density-based clustering (DBSCAN) to capture non-spherical groups, add bootstrapped stability analysis, or enrich the dataset with sensory scores to relate chemistry to perceived quality.
>>>>>>> theirs
